Mergeable Types for Highly-Available Replicated Data

### 2017-02-25 paper/mergeable-types.tex
Mergeability Def. 2

\caption{Commutative diagram representing the merge operation.}
\label{fig:pushouts}
\end{figure}

@@

@@

\begin{definition} [\bfseries Mergeable Type]
\label{def:mergeable-type-2}
A type $t$ is said to be mergeable if and only if the following
conditions hold:
\begin{itemize}
  \item There exists a function $W: t \times t \rightarrow (t
  \rightarrow t)\; \C{list}$ that accepts an object $A:t$ and its
  semantic successor $B:t$, and returns a minimal sequence of
  primitive morphisms $P^*$, whose composition maps $A$ to $B$. That
  is, $W(A,B) = P^*$, where $\denot{P^*}(A) = B$, and there does not
  exist an $R^*$ such that $\denot{R^*}(A) = B$ and $\length{R^*} <
  \length{P^*}$

  \item There exists a function $M: (t \rightarrow t)^*\!\times\!(t
  \rightarrow t)^* \;\rightarrow\; (t \rightarrow t)\!\times\!(t
  \rightarrow t)$ that accepts a pair of minimal sequences of
  primitive morphisms, $P^*$ and $Q^*$, and returns a pair of morphisms,
  $Q'$ and $P'$, such that for any object $A:t$, $(Q' \circ
  \denot{P^*})(A) = (P' \circ \denot{Q^*})(A)$.  
\end{itemize}
\end{definition}

\paragraph{Strings} Strings are mergeable. Given a string ($s$) and
its semantic successor ($s'$), an edit script ($P^*$ or $Q^*$) is the
shortest sequence of \C{insert}, \C{delete}, and \C{substitute}
operations (string's primitive morphisms), whose composition takes $s$
to $s'$. The length of the sequence is in fact the Levenshtein
distance~\cite{levenstein} between $s$ and $s'$, which can be computed
in polynomial time via Wagner-Fischer algorithm~\cite{wagner-fischer}.
A slight modification of Wagner-Fischer also lets us compute the edit
script. Fig.~\ref{fig:mergeable-string-eg} shows the edit script ...

>>>>

\caption{Commutative diagram for Def.~\ref{def:mergeable-type}.}
\label{fig:pushouts}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.4]{Figures/pushouts-2}

\caption{Commutative diagram for Def.~\ref{def:mergeable-type-2}.}
\label{fig:pushouts-2}
\end{figure}

@@

\paragraph{Notation} As mentioned earlier, we distinguish between the
named functions defined by a data type library (primitive morphisms),
and their compositions (morphisms). A sequence (list) of primitive
morphisms is written $P^*$ or $Q^*$. The composition of a sequence of
primitive morphisms (written $\denot{P^*}$) is their right-to-left
composition, i.e., $\C{fold\_left}\;(\lambda f.\lambda g. f \circ
g)\; id\; P^*$, where $\C{fold\_left}$ has its usual OCaml type. The
length of a sequence $S$ is written $\length{S}$.

\begin{definition} [\bfseries Mergeable Type]
\label{def:mergeable-type-2}
A type $t$ is said to be mergeable if and only if the following
conditions hold:
\begin{itemize}
  \item There exists a function $W: t \times t \rightarrow (t
  \rightarrow t)^*$ that accepts an object $A:t$ and its
  semantic successor $B:t$, and returns a minimal sequence of
  primitive morphisms $P^*$, whose composition maps $A$ to $B$. That
  is, $W(A,B) = P^*$, where $\denot{P^*}(A) = B$, and there does not
  exist an $R^*$ such that $\denot{R^*}(A) = B$ and $\length{R^*} <
  \length{P^*}$

  \item There exists a function $M: (t \rightarrow t)^*\!\times\!(t
  \rightarrow t)^* \;\rightarrow\; (t \rightarrow t)\!\times\!(t
  \rightarrow t)$ that accepts a pair of minimal sequences of
  primitive morphisms, $P^*$ and $Q^*$, and returns a pair of morphisms,
  $P'$ and $Q'$, such that for any object $A:t$, $(Q' \circ
  \denot{P^*})(A) = (P' \circ \denot{Q^*})(A)$.  
\end{itemize}
Commutativity diagram of Fig.~\ref{fig:pushouts-2} visualizes this
definition. 
\end{definition}

The definition asserts the existence of two functions for a mergeable
type $t$. First, the function $W$, which computes the \emph{edit
distance} between an object $A:t$ and its semantic successor $B:t$.
The sequence of primitive morphisms it returns ($P^*$) is called an
\emph{edit script}. Replaying the edit script on $A$ produces $B$
(i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the
existence of a function $M$, which performs an \emph{operational
transformation} of the primitive morphisms in an edit script. An
operational transformation of a primitive morphism $P$ w.r.t a
morphism $Q$ is another primitive morphism $P'$ such that for any
object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
$A$. We will now consider examples of mergeable types that reify the
notions of edit scripts and operational transformations.

\begin{figure}
\centering
\includegraphics[scale=0.4]{Figures/string-eg}

\caption{Merging strings with edit scripts and operational
transformations}
\label{fig:string-eg}
\end{figure}

\paragraph{Strings} Strings are mergeable. Given a string ($s$) and
its semantic successor ($s'$), an edit script ($P^*$ or $Q^*$) is the
shortest sequence of \C{insert}, \C{delete}, and \C{substitute}
operations (string's primitive morphisms), whose composition takes $s$
to $s'$. The length of the sequence is in fact the Levenshtein
distance~\cite{levenstein} between $s$ and $s'$, which can be computed
in polynomial time via Wagner-Fischer algorithm~\cite{wagner-fischer}.
A slight modification of Wagner-Fischer also lets us compute the edit
script. Fig.~\ref{fig:string-eg} illustrates.
